---
title: "EDA"
author: "Eric Song"
format: html
editor: visual
---


## Introduction 
This exploratory data analysis investigates the [CDC Diabetes Dataset] (https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/) to understand factors associated with diabetes prevalence. We want to find which biological and lifestyle factors can influence diabetes.  Primarily the variables we were interested in were:
  -If the participant has High Blood Pressure (Yes/No)
  -If the participant has High Cholesterol (Yes/No)
  -If the participant does physical activity (Yes/No)
  -The participant's BMI (Integer)
  -The participant's Age (Integer)
  -The participant's Gender (M/F)
  -The participants perception of their overall health     
  
These factors are analyzed to explore their relationships with the binary outcome variable, Diabetes_binary, which indicates the presence or absence of diabetes. By examining these variables, this analysis aims to uncover patterns and insights that may inform predictive modeling for diabetes risk.


```{r,warning=FALSE}
#Loading Libraries
library(tidyverse)
library(tidymodels)
```

```{r}
#Loading our data.
beetus_data<-read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")
#Changing to factors and adding names.  
beetus_data <- beetus_data |>
  mutate(
    Diabetes_binary = factor(Diabetes_binary, labels = c("No Diabetes", "Pre/Diabetes")),
    HighBP = factor(HighBP, labels = c("No High BP", "High BP")),
    HighChol = factor(HighChol, labels = c("No High Cholesterol", "High Cholesterol")),
    Smoker = factor(Smoker, labels = c("Non-Smoker", "Smoker")),
    PhysActivity = factor(PhysActivity, labels = c("No Physical Activity", "Physical Activity")),

    Sex = factor(Sex, labels = c("Female", "Male")),
    GenHlth = factor(GenHlth, levels=1:5, labels = c("Excellent", "Very Good", "Good", "Fair", "Poor")),
    Education = factor(Education,levels=1:6, labels = c(
      "Never Attended School or only kindergarten",
      "Grades 1-8",
      "Grades 9-11",
      "Grade 12/GED",
      "Some College/Technical School",
      "College Graduate"
    )),
    Income = factor(Income,levels=1:8, labels = c(
      "Less than $10,000",
      "$10,000 to $15,000",
      "$15,000 to $20,000",
      "$20,000 to $25,000",
      "$25,000 to $35,000",
      "$35,000 to $50,000",
      "$50,000 to $75,000",
      "$75,000 or more"
    )),
    Age = factor(Age,levels=1:13, labels = c(
      "18-24", "25-29", "30-34", "35-39", "40-44",
      "45-49", "50-54", "55-59", "60-64", "65-69",
      "70-74", "75-79", "80 or older"
    ))
  )
#Variables we're interested in
beetus_filtered <- beetus_data |>
  select(Diabetes_binary,HighBP,HighChol,PhysActivity,BMI,Age,Sex,GenHlth)

```
Let's take a look at the structure of our data and check if there's any missing values.


```{r}
#Quick look at data types
str(beetus_filtered)
#Checking for Missing Data
beetus_filtered |>
  is.na()|>
  colSums()
```

No missing values.  Let's check some numerical summaries for BMI.


## Numeric Variables

```{r}
beetus_filtered |>
  select(where(is.numeric)) |>
  summary()
```
BMI seems slightly skewed.  Let's graph it vs our diabetes binary variable.

```{r}
ggplot(beetus_filtered, aes(x = BMI, fill = Diabetes_binary)) +
  geom_bar() 
ggplot(beetus_filtered, aes(x = Diabetes_binary,y = BMI, fill = Diabetes_binary)) +
  geom_violin() +
  labs(
    title = "Violin Plot of BMI by Diabetes Status",
    x = "Diabetes Status",
    y = "BMI",
    fill = "Diabetes Status"
  )
```
On visual inspection, there does seem to be a difference in our samples.  
Now let's look at our categorical variables with contingency tables.

## Categorical Variables

```{r}
cont_tables <- lapply(names(beetus_filtered)[-which(names(beetus_filtered) == "Diabetes_binary")], function(var) {
  table(beetus_filtered[[var]], beetus_filtered$Diabetes_binary)
})

names(cont_tables) <- names(beetus_filtered)[-which(names(beetus_filtered) == "Diabetes_binary")]

cont_tables


```

From the contingency table for HighBP, we observe a higher proportion of individuals with diabetes who have high blood pressure compared to those without diabetes. This reinforces the connection between hypertension and diabetes.

Let's make some graphs now.

```{r}
#Showing plots of variables with 2 factor levels
twofact_plot<- beetus_filtered |>
  #Select variables with only 2 factor levels for clean plotting
  select(Diabetes_binary, where(~ is.factor(.) && nlevels(.) == 2)) |>
  pivot_longer(
    cols = -Diabetes_binary, 
    names_to = "Variable",   
    values_to = "Value"      
  )
#some scaling issues
ggplot(twofact_plot, aes(x = Value, fill = Diabetes_binary)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ Variable, scales = "free_x") + 
  #tilted for readability
  theme(axis.text.x = element_text(angle = 10))

#Plots for more than 2 factor levels
multifact_data <- beetus_filtered |>
  select(Diabetes_binary, where(~ is.factor(.) && nlevels(.) > 2)) |>
  pivot_longer(
    cols = -Diabetes_binary, 
    names_to = "Variable",   
    values_to = "Value"      
  )
# Create individual plots for each variable
plots <- list() # Empty list to store plots
# Get unique variable names
unique_variables <- unique(multifact_data$Variable) 
for (var in unique_variables) {
  plot <- multifact_data |>
    filter(Variable == var) |> 
    ggplot(aes(x = Value, fill = Diabetes_binary)) +
    geom_bar(position = "dodge") +
    labs(
      title = paste("Distribution of", var, "by Diabetes Status"),
      x = var,
      y = "Count",
      fill = "Diabetes Status"
    ) +
    
    theme(axis.text.x = element_text(angle = 10))
  # Store the plot in the list
  plots[[var]] <- plot
}
plots[[unique_variables[1]]]
plots[[unique_variables[2]]]
```
General health seems to have a visual difference as well.

### Conclusion

This analysis highlights that BMI, HighBP, and self-reported health ratings (GenHlth) appear to have differences. These variables could serve as valuable inputs for predictive modeling. Our modeling section will further explore these variables to see if they are significant.


[Click here for the Modeling Page](Modeling.html)
