---
title: "EDA"
author: "Eric Song"
format: html
editor: visual
---

## Exploration

Let's take a look at our data.  Let's start with looking at the data/variable types and checking for missing values.

```{r}
library(tidyverse)
library(tidymodels)
beetus_data<-read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")
#Quick look at data types
str(beetus_data)
#Checking for Missing Data
beetus_data |>
  is.na()|>
  colSums()
```

There seems to be no missing values.  Based on the look at structure, we might have some factor variables.  Before that though, checking a quick summary for our variables. 

```{r}
summary(beetus_data)
```

Looks like we have some factors.  Checking for unique values.

```{r}
beetus_data |>
  summarise(across(everything(), ~ list(unique(.))))
```

Looks like we do have multiple factors.  <dbl[2]>s are binary values.  Education/income seem to be factors as well.  Going to the dataset's source website: https://archive.ics.uci.edu/dataset/891/cdc+diabetes+health+indicators
we find descriptions for our variables.  Applying formatting to our variables now.

```{r}
beetus_data <- beetus_data |>
  mutate(
    Diabetes_binary = factor(Diabetes_binary, levels=c('0','1'), labels = c("No Diabetes", "Pre/Diabetes")),
    HighBP = factor(HighBP, labels = c("No High BP", "High BP")),
    HighChol = factor(HighChol, labels = c("No High Cholesterol", "High Cholesterol")),
    CholCheck = factor(CholCheck, labels = c("No Cholesterol Check", "Cholesterol Check")),
    Smoker = factor(Smoker, labels = c("Non-Smoker", "Smoker")),
    Stroke = factor(Stroke, labels = c("No Stroke", "Stroke")),
    HeartDiseaseorAttack = factor(HeartDiseaseorAttack, labels = c("No", "Yes")),
    PhysActivity = factor(PhysActivity, labels = c("No Physical Activity", "Physical Activity")),
    Fruits = factor(Fruits, labels = c("No", "Yes")),
    Veggies = factor(Veggies, labels = c("No", "Yes")),
    HvyAlcoholConsump = factor(HvyAlcoholConsump, labels = c("No", "Yes")),
    AnyHealthcare = factor(AnyHealthcare, labels = c("No", "Yes")),
    NoDocbcCost = factor(NoDocbcCost, labels = c("No", "Yes")),
    DiffWalk = factor(DiffWalk, labels = c("No", "Yes")),
    Sex = factor(Sex, labels = c("Female", "Male")),
    GenHlth = factor(GenHlth, labels = c("Excellent", "Very Good", "Good", "Fair", "Poor")),
    Education = factor(Education, labels = c(
      "Never Attended School or only kindergarten",
      "Grades 1-8",
      "Grades 9-11",
      "Grade 12/GED",
      "Some College/Technical School",
      "College Graduate"
    )),
    Income = factor(Income, labels = c(
      "Less than $10,000",
      "$10,000 to $15,000",
      "$15,000 to $20,000",
      "$20,000 to $25,000",
      "$25,000 to $35,000",
      "$35,000 to $50,000",
      "$50,000 to $75,000",
      "$75,000 or more"
    )),
    Age = factor(Age, labels = c(
      "18-24", "25-29", "30-34", "35-39", "40-44",
      "45-49", "50-54", "55-59", "60-64", "65-69",
      "70-74", "75-79", "80 or older"
    ))
  )
str(beetus_data)
```

18 Categorical Variables and 3 Numerical variables.  Now our data is tidy,
let's make some contingency tables since we have a bunch of factors.
```{r}
cont_tables <- lapply(names(beetus_data)[-which(names(beetus_data) == "Diabetes_binary")], function(var) {
  table(beetus_data[[var]], beetus_data$Diabetes_binary)
})

names(cont_tables) <- names(beetus_data)[-which(names(beetus_data) == "Diabetes_binary")]

cont_tables
```


Let's make some graphs as well.

#distributions for bmi/age etc...
```{r}
ggplot(beetus_data, aes(x = Diabetes_binary)) +
  geom_bar()

ggplot(beetus_data, aes(x =  Diabetes_binary )) + geom_bar()
ggplot(beetus_data, aes(x =  Smoker )) + geom_bar()
ggplot(beetus_data, aes(x =  HighBP )) + geom_bar()
ggplot(beetus_data, aes(x =  HighChol )) + geom_bar()
```
```{r}
```

```{r}
#Correlation Matrix
beetus_data |>
select(where(is.numeric)) |>
cor() |>
round(3)
```

[Click here for the Modeling Page](Modeling.html)
